---
title: 'ESM 263: Assignment 4'
author: "Linus Blomqvist"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
---

```{r setup, include = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

```{r}
# install.packages("pacman") -- run this if you don't already have pacman on your machine
pacman::p_load(sf, raster, tmap, tidyverse, units, scales, rgdal, rgeos, tmaptools,
               tictoc, kableExtra)
```

# Intro and caveat
There appears to be no package in R that can do viewshed analysis. There is a [package](https://www.rdocumentation.org/packages/windfarmGA/versions/2.3.0/topics/viewshed) that does something along those lines but from what I can tell it doesn't perfectly suit my purposes. The only way to do the viewshed analysis is by drawing on QGIS (there is a mention of this [here](https://geocompr.robinlovelace.net/gis.html#gis)). But that goes a bit beyond what I have time to figure out so I'm just going to use two of the criteria - develop and riparian - and create a weighted score for each watershed based on those two. The original weights were 0.50 for riparian and 0.17 for developable. To preserve the ratio between these two, I will assign weights of 0.75 for riparian and 0.25 for developable.

## Load data

We have the following data at our disposal, sorted under the respective data files:

* DEM (raster)
* Landsat 7 (raster)
* Input (geopackage)
  + Watersheds (vector)
  + Streams (vector)
  + Parcels (vector)
  + ROI (vector)
  + Observation points (vector)

Let's first check the input file.
```{r}
(input_layers <- st_layers("HW4/data_hw4/input.gpkg")$name)
```

Reading these in below. Note: since `tmap` no longer works with sf objects of class tibble, I will read these files in with `st_read` instead of `read_sf`.

```{r read_gpkg, results = "hide"}
parcels <- st_read("HW4/data_hw4/input.gpkg", layer = input_layers[1])
obs_points <- st_read("HW4/data_hw4/input.gpkg", layer = input_layers[2])
ROI <- st_read("HW4/data_hw4/input.gpkg", layer = input_layers[3])
streams <- st_read("HW4/data_hw4/input.gpkg", layer = input_layers[4])
watersheds <- st_read("HW4/data_hw4/input.gpkg", layer = input_layers[5])
```

I'm going to plot all these to see what they look like. Here are the parcels:

```{r}
tm_shape(parcels) +
  tm_polygons(alpha = 0.5) +
  tm_layout(main.title = "Parcels",
            frame = FALSE, 
            main.title.size = 1.5,
            main.title.position = "center")
```

And the Region of Interest and the observation points:

```{r}
tm_shape(ROI) +
  tm_borders(lwd = 1.5) +
  tm_shape(obs_points) +
  tm_dots(col = "blue", size = 0.1) +
  tm_layout(main.title = "ROI and observation points",
            frame = FALSE, 
            main.title.size = 1.5,
            main.title.position = "center") +
  tm_add_legend(type = "symbol", col = "blue", labels = "Observation points") +
  tm_add_legend(type = "line", col = "black", labels = "ROI: Santa Barbara County")
```

And the watersheds and streams:

```{r}
tm_shape(ROI) +
  tm_borders(col = "black") +
  tm_shape(watersheds) +
  tm_borders(col = "darkgreen", alpha = 0.8, lwd = 1.5) +
  tm_shape(streams) +
  tm_lines(col = "blue", alpha = 0.3) +
  tm_layout(main.title = "Watersheds and streams",
            frame = FALSE, 
            main.title.size = 1.5,
            main.title.position = "center") +
  tm_add_legend(type = "line", col = "darkgreen", labels = "Watershed boundaries") +
  tm_add_legend(type = "line", col = "blue", labels = "Streams")
```

Let's check the CRS:

```{r}
st_crs(parcels)$input
```

Now I'll read in the DEM and clip (mask) it to the ROI for convenience.

```{r}
DEM <- raster("HW4/data_hw4/dem100.tif")
DEM <- mask(DEM, ROI)
```

Let's check the CRS:

```{r}
crs(DEM)
```

This looks the same as the CRS for the watershed data:

```{r}
st_crs(watersheds)$proj4string
```

Now I plot the DEM: 

```{r}
tm_shape(DEM) +
  tm_raster(title = "Elevation") +
  tm_shape(ROI) +
  tm_borders() +
  tm_legend(position = c("left", "bottom"), 
            legend.bg.color = "white", 
            legend.bg.alpha = 0.7)
```

Read in Landsat data and check names of layers:

```{r}
landsat <- stack("HW4/data_hw4/Landsat71.tif")
names(landsat)
```

And plot the bands

```{r}
tmap_mode("plot")
tm_shape(landsat) +
  tm_raster()
```


With that, we've read in all the data.

## Riparian criterion

Here, the task is to calculate the percentage of each watershed that falls within 1000 ft of a stream and where NDVI $\in [0.2,0.5]$. The way I do this is as follows. I create a buffer around streams. I then calculate NDVI for all pixels, and make a raster "dummy" for the appropriate NDVI range. By masking this dummy raster to the stream buffer zones, I am left with a raster with 1s for pixels that have the right NDVI and are within 1000 ft of streams, and 0s for all other pixels. I then add up the number of such pixels for each watershed and create a normalized score.

```{r}

# Create a 1000 ft (304.8 m) buffer around streams
stream_buffer <- st_buffer(streams, 304.8)

# Calculate NDVI
NDVI <- (landsat$Landsat71.4 - landsat$Landsat71.3) / (landsat$Landsat71.4 + landsat$Landsat71.3)

# Dummy for pixels with the requisite NDVI
suitable_NDVI <- NDVI >= 0.2 & NDVI <= 0.5

# Limit the NDVI raster to the stream buffer zones
NDVI_and_riparian <- mask(x = suitable_NDVI, mask = stream_buffer)

# Calculate the number of pixels meeting both criteria, by watershed
watersheds$riparian <- NDVI_and_riparian %>%
  raster::extract(watersheds, fun = sum, na.rm = TRUE) %>%
  as.vector()

# Normalize
watersheds <- mutate(watersheds, z_score_riparian = (riparian - mean(riparian)) / sd(riparian))

```

## Develop criterion

My reasoning for the develop criterion is as follows. You can't buy part of a parcel, you have to buy the whole thing. So if a parcel has any part that is developable, you need to buy the whole parcel to prevent that development. Below, I first identify the pixels with slope less than 20%. I then deem developable a parcel that has at least one developable pixel and that is not public. By attaching the objectid of the watersheds to each parcel, I can group the parcels by watershed and sum up the cost of all the developable parcels in each watershed. I then join this to the watershed dataframe, and create a normalized score for each watershed.

```{r}

# Create slope raster
slope <- terrain(DEM, opt = "slope", unit = "radians", neighbors = 8)

# Create dummies for those pixels with a slope less than 20%
# Note: A 20% slope corresponds to `r atan(0.02)` radians.
slope_sub_20 <- slope < atan(0.2)

# Add up the number of developable pixels in each parcel
parcels$suitable_slope <- slope_sub_20 %>%
  raster::extract(parcels, fun = sum, na.rm = TRUE) %>%
  as.vector()

# If there is one or more developable pixels in a parcel, one has to buy it 
# to prevent development, so I create a dummy for developable
# In this dummy I also exclude all public parcels
parcels$developable <- parcels$suitable_slope > 0 & is.na(parcels$NONTAXCODE)

# Attach the objectid of watersheds to each parcel
parcels <- st_join(parcels, watersheds["objectid"])

# Calculate total purchasing cost by watershed
total_purchase_cost <- parcels %>%
  st_drop_geometry() %>%
  filter(developable == 1) %>% # non-developable parcels cost nothing
  group_by(objectid) %>% # objectid of the watersheds
  summarize(tot_purchase_cost = sum(NET_AV))

# Join on this purchasing cost to the watershed data
watersheds <- left_join(watersheds, total_purchase_cost, by = "objectid")

# Those that have NA have no developable land so the cost is 0
watersheds$tot_purchase_cost[is.na(watersheds$tot_purchase_cost)] <- 0

# Create z score
watersheds <- mutate(watersheds, z_score_develop = (tot_purchase_cost - mean(tot_purchase_cost, na.rm = TRUE)) / sd(tot_purchase_cost, na.rm = TRUE))

```

# MCA

Each watershed now gets a total weighted score, as follows. I take the negative of the purchase cost since a higher value is worse.

```{r}
watersheds$final_score <- (watersheds$z_score_riparian * 0.75) + 
  (-watersheds$z_score_develop * 0.25)
```


```{r}
score_table <- watersheds %>%
  select(objectid, cdfspwname, final_score) %>%
  arrange(desc(final_score)) %>%
  st_drop_geometry()

kbl(score_table, caption = "Table 1: Final scores",
      col.names = c("Object ID", "Watershed Name", "Final score")) %>%
  kable_styling(bootstrap_options = "striped",
                full_width = FALSE)
```

```{r}
tmap_mode("plot")
tm_shape(watersheds) +
  tm_fill(col = "final_score",
          title = "Final score",
          palette = "Spectral") +
  tm_layout(main.title = "Final scores for watersheds",
            main.title.position = c("center", "top"),
            legend.bg.color = "white") +
  tm_credits("The final score for each watershed is the weighted sum of the 
normalized scores for the amount of riparian land in the watershed 
and the (negative of) the cost of avoiding development.",
             position = c("center", "bottom"),
              bg.color = "white",
              bg.alpha = 0.7)
```

